<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Binary segmentation examples}
-->

# Binary segmentation examples

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this vignette we explore the differences in asymptotic time
complexity between different implementations of binary segmentation.

## Asymptotic time/memory measurement of different R expressions

The code below uses the following arguments:

* `N` is a numeric vector of data sizes,
* `setup` is an R expression to create the data,
* the other arguments have names to identify them in the results, and
  values which are R expressions to time,

```{r}
library(data.table)
atime.list <- atime::atime(
  N=2^seq(2, 20),
  setup={
    max.segs <- as.integer(N/2)
    max.changes <- max.segs-1L
    set.seed(1)
    data.vec <- 1:N
  },
  "changepoint::cpt.mean"={
    cpt.fit <- changepoint::cpt.mean(data.vec, method="BinSeg", Q=max.changes)
    sort(c(N,cpt.fit@cpts.full[max.changes,]))
  },
  "binsegRcpp::binseg_normal"={
    binseg.fit <- binsegRcpp::binseg_normal(data.vec, max.segs)
    sort(binseg.fit$splits$end)
  },
  "fpop::multiBinSeg"={
    mbs.fit <- fpop::multiBinSeg(data.vec, max.changes)
    sort(c(mbs.fit$t.est, N))
  },
  "wbs::sbs"={
    wbs.fit <- wbs::sbs(data.vec)
    split.dt <- data.table(wbs.fit$res)[order(-min.th, scale)]
    sort(split.dt[, c(N, cpt)][1:max.segs])
  },
  binsegRcpp.list={
    binseg.fit <- binsegRcpp::binseg(
      "mean_norm", data.vec, max.segs, container.str="list")
    sort(binseg.fit$splits$end)
  },
  ##seconds.limit=0.1,
  times=5)
plot(atime.list)
```

The default plot method creates a log-log plot of median time vs data
size, for each of the specified R expressions. You can use
`references_best` to get a tall/long data table that can be plotted to
show both empirical time and memory complexity:

```{r}
best.list <- atime::references_best(atime.list)
if(require(ggplot2)){
  hline.df <- with(atime.list, data.frame(seconds.limit, unit="seconds"))
  gg.both <- ggplot()+
    theme_bw()+
    facet_grid(unit ~ ., scales="free")+
    geom_hline(aes(
      yintercept=seconds.limit),
      color="grey",
      data=hline.df)+
    geom_line(aes(
      N, empirical, color=expr.name),
      data=best.list$meas)+
    geom_ribbon(aes(
      N, ymin=min, ymax=max, fill=expr.name),
      data=best.list$meas[unit=="seconds"],
      alpha=0.5)+
    scale_x_log10()+
    scale_y_log10("median line, min/max band")
  if(require(directlabels)){
    gg.both+
      directlabels::geom_dl(aes(
        N, empirical, color=expr.name, label=expr.name),
        method="right.polygons",
        data=best.list$meas)+
      theme(legend.position="none")+
      coord_cartesian(xlim=c(1,2e7))
  }else{
    gg.both
  }
}
```

The plots above show some speed differences between binary
segmentation algorithms, but they could be even easier to see for
larger data sizes (exercise for the reader: try modifying the `N` and
`seconds.limit` arguments). You can also see that memory usage is much
larger for changepoint than for the other packages.

## Closest asymptotic references

You can use code like below to compute asymptotic references which are
best fit for each expression. We do the best fit by adjusting each
reference to the largest N, and then ranking each reference by
distance to the measurement of the second to largest N. The code below
uses `each.sign.rank==1` to compute the closest reference above and
below,

```{r}
best.refs <- best.list$ref[each.sign.rank==1]
(time.refs <- best.refs[unit=="seconds"])
```

Then you can plot these references with the empirical data using the
ggplot code below,

```{r}
ref.color <- "red"
## try() to avoid CRAN error 'from' must be a finite number, on
## Flavors: r-devel-linux-x86_64-debian-gcc, r-release-linux-x86_64,
## due to https://github.com/r-lib/scales/issues/307
(seconds.dt <- best.list$meas[unit=="seconds"])
try(if(require(ggplot2)){
  gg <- ggplot()+
    geom_line(aes(
      N, reference, group=fun.name),
      color=ref.color,
      data=time.refs)+
    geom_line(aes(
      N, empirical),
      linewidth=1,
      data=seconds.dt)+
    scale_x_log10()+
    scale_y_log10("median line, min/max band")+
    facet_wrap("expr.name")+
    theme_bw()
  if(require(directlabels)){
    gg+
      directlabels::geom_dl(aes(
        N, reference, label=fun.name),
        data=time.refs,
        color=ref.color,
        method="bottom.polygons")
  }else{
    gg
  }
})
```

## Custom asymptotic references

If you have one or more expected time complexity classes that you want
to compare with your empirical measurements, you can use the
`fun.list` argument:

```{r}
my.refs <- list(
  "N \\log N"=function(N)log10(N) + log10(log(N)),
  "N^2"=function(N)2*log10(N),
  "N^3"=function(N)3*log10(N))
my.best <- atime::references_best(atime.list, fun.list=my.refs)
```

Note that in the code above, each R function should take as input the
data size `N` and output log base 10 of the reference function.

```{r}
(my.best.time.refs <- my.best$ref[unit=="seconds"])
try(if(require(ggplot2)){
  gg <- ggplot()+
    geom_line(aes(
      N, reference, group=fun.name),
      color=ref.color,
      data=my.best.time.refs)+
    geom_line(aes(
      N, empirical),
      linewidth=1,
      data=seconds.dt)+
    scale_x_log10()+
    scale_y_log10("median line, min/max band")+
    facet_wrap("expr.name")+
    theme_bw()
  if(require(directlabels)){
    gg+
      directlabels::geom_dl(aes(
        N, reference, label=fun.name),
        data=my.best.time.refs,
        color=ref.color,
        method="bottom.polygons")
  }else{
    gg
  }
})
```

From the plot above you should be able to see the asymptotic time
complexity class of each algorithm.

* the fastest packages are log-linear (fpop, wbs, binsegRcpp), 
* that the binsegRcpp implementation that uses the list container is
  much slower (quadratic).
* the changepoint package uses a super-quadratic algorithm (actually
  cubic, which you could see if you increase N even larger).

## Exercises for the reader

* increase `seconds.limit` to see the differences more clearly.
* compute and plot asymptotic references for memory instead of time
  (including expected memory complexity, linear).
