<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Comparing git versions of data.table}
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

In this vignette we show you how to compare asymptotic timings of an R
expression which uses different versions of the `data.table` package,
which we clone from github using the code below,

```{r}
dt.git2r <- tempfile()
dir.create(dt.git2r)
dt.gh.url <- "https://github.com/tdhock/data.table"
git2r::clone(dt.gh.url, dt.git2r)
```

After cloning the git repo, we give the repo path as the `pkg.path`
argument to `atime_versions` in the code below. We also need the
following arguments:

* `N` is the sequence of data sizes,
* `setup` is an expression that will be evaluated for every data size,
  prior to measuring time/memory,
* `expr` is an expression that will be evaluated for all of the
  different git commit versions. It must call a function from the
  cloned package, using double or triple colon prefix (the package
  named before the colons will be replaced by a new package name that
  uses the commit SHA hash). Below we use `data.table:::[.data.table`,
  which will become something like
  `data.table.3fa8b20435d33b3d4b5c26fd9b0ac14c10b98800:::[.data.table`
  for each of the different package versions.

In the code below we also use `pkg.edit.fun` to specify how the
package must be edited so that it can install and load using a
version-specific package name `PKG.SHA`. It is not necessary to
specify `pkg.edit.fun` for typical packages (no compiled code, or
Rcpp), but `data.table` is an interesting example use case for
`pkg.edit.fun` since it specifies a custom shared object file name in
`Makevars`, and it has some custom version checking code in
`onLoad.R`.

The other arguments in the code below have names which identify the
different versions of the code, and values which are commit SHA
hashes. The particular commits chosen below were recommended by [git
bisect](https://git-scm.com/docs/git-bisect), and the expressions were
adapted from issue
[#5424](https://github.com/Rdatatable/data.table/pull/5424).

```{r}
run.atime <- function(TDIR){
  atime::atime_versions(
    pkg.path=TDIR,
    pkg.edit.fun=function(old.Package, new.Package, sha, new.pkg.path){
      pkg_find_replace <- function(glob, FIND, REPLACE){
        atime::glob_find_replace(file.path(new.pkg.path, glob), FIND, REPLACE)
      }
      Package_regex <- gsub(".", "_?", old.Package, fixed=TRUE)
      Package_ <- gsub(".", "_", old.Package, fixed=TRUE)
      new.Package_ <- paste0(Package_, "_", sha)
      pkg_find_replace(
        "DESCRIPTION", 
        paste0("Package:\\s+", old.Package),
        paste("Package:", new.Package))
      pkg_find_replace(
        file.path("src","Makevars.*in"),
        Package_regex,
        new.Package_)
      pkg_find_replace(
        file.path("R", "onLoad.R"),
        Package_regex,
        new.Package_)
      pkg_find_replace(
        file.path("R", "onLoad.R"),
        sprintf('packageVersion\\("%s"\\)', old.Package),
        sprintf('packageVersion\\("%s"\\)', new.Package))
      pkg_find_replace(
        file.path("src", "init.c"),
        paste0("R_init_", Package_regex),
        paste0("R_init_", gsub("[.]", "_", new.Package_)))
      pkg_find_replace(
        "NAMESPACE",
        sprintf('useDynLib\\("?%s"?', Package_regex),
        paste0('useDynLib(', new.Package_))
    },
    N = 10^seq(3, 8),
    setup={
      n <- N/100
      set.seed(1L)
      dt <- data.table(
        g = sample(seq_len(n), N, TRUE),
        x = runif(N),
        key = "g")
    },
    expr={
      dt_mod <- copy(dt)
      data.table:::`[.data.table`(dt_mod, , N := .N, by = g)
    },
    verbose = TRUE,
    "1.14.0 on CRAN. Bump to 1.14.1"="263b53e50241914a22f7ba6a139b52162c9d7927",
    "1.14.3 dev master"="c4a2085e35689a108d67dacb2f8261e4964d7e12",
    "1.14.4"="8fd1c8e0250fdd7eb1bad9f03678755071eeb087")
}
atime.list <- if(requireNamespace("callr")){
  requireNamespace("atime")
  callr::r(run.atime, list(dt.git2r))
}else{
  run.atime(dt.git2r)
}
```

The results can be plotted using the code below,

```{r}
best.list <- atime::references_best(atime.list)
both.dt <- best.list$meas
if(require(ggplot2)){
  hline.df <- with(atime.list, data.frame(seconds.limit, unit="seconds"))
  gg <- ggplot()+
    theme_bw()+
    facet_grid(unit ~ ., scales="free")+
    geom_hline(aes(
      yintercept=seconds.limit),
      color="grey",
      data=hline.df)+
    geom_line(aes(
      N, empirical, color=expr.name),
      data=best.list$meas)+
    geom_ribbon(aes(
      N, ymin=min, ymax=max, fill=expr.name),
      data=best.list$meas[unit=="seconds"],
      alpha=0.5)+
    scale_x_log10()+
    scale_y_log10("median line, min/max band")
  if(require(directlabels)){
    gg+
      directlabels::geom_dl(aes(
        N, empirical, color=expr.name, label=expr.name),
        method="right.polygons",
        data=best.list$meas)+
      theme(legend.position="none")+
      coord_cartesian(xlim=c(1e3,1e10))
  }else{
    gg
  }
}
```

The figure above shows that there are significant differences between
the timings of the commits. 

* version 1.14.0 on CRAN is fast (relatively
  small computation time),
* version 1.14.3 dev master is slow (relatively large computation time)
* version 1.14.4 is fast again (bug fixed),
* a similar analyis was used in https://github.com/Rdatatable/data.table/issues/5371 to fix the bug.

Below we remove the installed packages,

```{r}
atime::atime_versions_remove("data.table")
```
