<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Comparing git versions}
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

# Comparing asymptotic timings of git versions

In this vignette we explain how to use functions which compute
asymptotic timings of different git versions of a package (useful for
determining when a difference in performance started to happen).

## Basic usage, `atime_versions` function

In this vignette we show you how to compare asymptotic timings of an R
expression which uses different versions of a package. Let us begin by
cloning the binsegRcpp package,

```{r}
old.opt <- options(width=100)
pkg.path <- tempfile()
dir.create(pkg.path)
git2r::clone("https://github.com/tdhock/binsegRcpp", pkg.path)
```

Next, to satisfy the CRAN requirement that we can not install packages
to the default library, we must create a library under /tmp,

```{r}
tmp.lib.path <- tempfile()
dir.create(tmp.lib.path)
lib.path.vec <- c(tmp.lib.path, .libPaths())
.libPaths(lib.path.vec)
```

Next, we define a helper function `run.atime` that will run
`atime_versions`, which is a simple way to compare different github
versions of a function:

```{r}
run.atime.versions <- function(PKG.PATH, LIB.PATH){
  if(!missing(LIB.PATH)).libPaths(LIB.PATH)
  atime::atime_versions(
    pkg.path=PKG.PATH,
    N=2^seq(2, 20),
    setup={
      max.segs <- as.integer(N/2)
      data.vec <- 1:N
    },
    expr=binsegRcpp::binseg_normal(data.vec, max.segs),
    cv="908b77c411bc7f4fcbcf53759245e738ae724c3e",
    "rm unord map"="dcd0808f52b0b9858352106cc7852e36d7f5b15d",
    "mvl_construct"="5942af606641428315b0e63c7da331c4cd44c091")
}
```

Here is an explanation of the arguments specified above:

* `pkg.path` is the path to the github repository containing the R package,
* `N` is a numeric vector of data sizes,
* `setup` is an R expression which will be run to create data for each size `N`,
* `expr` is an R expression which will be timed for each package
  version. Under the hood, a different R package is created for each
  package version, with package names like Package.SHA,
  `binsegRcpp.908b77c411bc7f4fcbcf53759245e738ae724c3e`. This `expr`
  must contain double or triple colon package name prefix code, like
  `binsegRcpp::binseg_normal` above, which will be translated to
  several different version-specific expressions, like
  `binsegRcpp.908b77c411bc7f4fcbcf53759245e738ae724c3e::binseg_normal`.
* The remaining arguments specify the different package versions
  (names for labels, values for SHA version IDs).

Note that in your code you don't have to create a helper function like
`run.atime.versions` in the code above. We do it in the package
vignette code, in order to run the different versions of the code
using `callr::r`, in a separate R process. This allows us to avoid
CRAN warnings about unexpected files found in the package check
directory, by safely deleting/removing the installed packages, after
having run the example code. For a more typical usage see
`example(atime_versions, package="atime")`.

In the code block below we compute the timings,

```{r}
atime.ver.list <- if(requireNamespace("callr")){
  requireNamespace("atime")
  callr::r(run.atime.versions, list(pkg.path, lib.path.vec))
}else{
  run.atime.versions(pkg.path)
}
names(atime.ver.list$measurements)
atime.ver.list$measurements[, .(N, expr.name, min, median, max, kilobytes)]
```

The result is a list with a `measurements` data table that contains
measurements of time in seconds (`min`, `median`, `max`) and memory
usage (`kilobytes`) for every version (`expr.name`) and data size
(`N`). A more convenient version of the data for plotting can be
obtained via the code below:

```{r}
plot(atime.ver.list)
```

## Advanced usage, `atime_versions_exprs` with `atime`

What if you wanted to compare different versions of one R package, to
another R package? Continuing the example above, we can get a list of
expressions, each one for a different version of the package, via the
code below:

```{r}
(ver.list <- atime::atime_versions_exprs(
  pkg.path=pkg.path,
  expr=binsegRcpp::binseg_normal(data.vec, max.segs),
  cv="908b77c411bc7f4fcbcf53759245e738ae724c3e",
  "rm unord map"="dcd0808f52b0b9858352106cc7852e36d7f5b15d",
  "mvl_construct"="5942af606641428315b0e63c7da331c4cd44c091"))
```

The `ver.list` created above can be augmented with other expressions,
such as the following alternative implementation of binary
segmentation from the changepoint package,

```{r}
expr.list <- c(ver.list, if(requireNamespace("changepoint")){
  list(changepoint=substitute(changepoint::cpt.mean(
    data.vec, penalty="Manual", pen.value=0, method="BinSeg",
    Q=max.segs-1)))
})
```

The `expr.list` created above can be provided as an argument to the
`atime` function as in the code below,

```{r}
run.atime <- function(ELIST, LIB.PATH){
  if(!missing(LIB.PATH)).libPaths(LIB.PATH)
  atime::atime(
    N=2^seq(2, 20),
    setup={
      max.segs <- as.integer(N/2)
      data.vec <- 1:N
    },
    expr.list=ELIST)
}
atime.list <- if(requireNamespace("callr")){
  requireNamespace("atime")
  callr::r(run.atime, list(expr.list, lib.path.vec))
}else{
  run.atime(expr.list)
}
```

Again note in the code above that we defined a helper function,
`run.atime`, and used `callr::r`, to avoid CRAN issues. For a more
typical usage see `example(atime_versions_exprs, package="atime")`.

```{r}
atime.list$measurements[, .(N, expr.name, median, kilobytes)]
```

The results above show that timings were computed for the three
different versions of the binsegRcpp code, along with the changepoint
code. These data can be plotted via the default method as in the code
below,

```{r}
refs.best <- atime::references_best(atime.list)
plot(refs.best)
```

## Cleanup

Below we remove the installed packages, in order to avoid CRAN warnings:

```{r}
atime::atime_versions_remove("binsegRcpp")
options(old.opt)
```
