<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Example: sparse matrices}
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
vec.mat.result <- atime::atime(
  N=10^seq(1,7,by=0.5),
  vector=numeric(N),
  matrix=matrix(0, N, N),
  Matrix=Matrix(0, N, N))
vec.mat.best <- atime::references_best(vec.mat.result)
plot(vec.mat.best)
```

The plot above shows that ICU/PCRE/TRE are all exponential in N
(subject/pattern size) when the pattern contains backreferences.

```{r}
all.exprs <- c(
  if(requireNamespace("re2"))atime::atime_grid(
    RE2=re2::re2_match(subject, pattern)),
  backtrackers)
all.result <- atime::atime(
  N=subject.size.vec,
  setup={
    subject <- paste(rep("a", N), collapse="")
    pattern <- paste(rep(c("a?", "a"), each=N), collapse="")
  },
  expr.list=all.exprs)
all.best <- atime::references_best(all.result)
plot(all.best)
```

The plot above shows that ICU/PCRE are exponential time whereas
RE2/TRE are polynomial time. Exercise for the reader: modify the above
code to use the `seconds.limit` argument so that you can see what
happens to ICU/PCRE for larger N (hint: you should see a difference at
larger sizes).

## Interpolate at seconds.limit using predict method

```{r}
(all.pred <- predict(all.best))
summary(all.pred)
```

The `predict` method above returns a list with a new element named
`prediction`, which shows the data sizes that can be computed with a
given time budget. The `plot` method is used below,

```{r}
plot(all.pred)
```

## `atime_grid` to compare different engines

In the `nc` package there is an `engine` argument which controls which
C regex library is used:

```{r}
nc.exprs <- atime::atime_grid(
  list(ENGINE=c(
    if(requireNamespace("re2"))"RE2",
    "PCRE",
    if(requireNamespace("stringi"))"ICU")),
  nc=nc::capture_first_vec(subject, pattern, engine=ENGINE))
nc.result <- atime::atime(
  N=subject.size.vec,
  setup={
    rep.collapse <- function(chr)paste(rep(chr, N), collapse="")
    subject <- rep.collapse("a")
    pattern <- list(maybe=rep.collapse("a?"), rep.collapse("a"))
  },
  expr.list=nc.exprs)
nc.best <- atime::references_best(nc.result)
plot(nc.best)
```

The result/plot above is consistent with the previous result.
